Trajectories

### 2022-08-27 docs/index.Rmd
Update index

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = TRUE,
  echo = FALSE,
  warning = TRUE,
  include = TRUE,
  out.width = "80%",
  dpi = 250
)
options(
  knitr.kable.NA = "-",
  knitr.duplicate.label = "allow",
  ggplot2.discrete.fill = wesanderson::wes_palettes$Darjeeling1,
  ggplot2.discrete.colour = wesanderson::wes_palettes$Darjeeling1,
  ggplot2.continuous.fill = ggplot2::scale_color_gradient,
  ggplot2.continuous.colour = ggplot2::scale_color_gradient
)

@@

@@

@@

```{r items_frequency, message=FALSE, warning=FALSE, paged.print=FALSE}

by_language <- items %>% 
  drop_na(frequency) %>% 
  ggplot(aes(frequency, lv, colour = language)) +
  geom_point(alpha = 0.5, shape = 1, stroke = 1) +
  labs(
    y = "Levenshtein similarity", 
    x = "Lexical frequency (Zipf score)\nExtracted from SUBTLEX", 
    colour = "Cognateness", 
    fill = "Cognateness"
  ) +
  scale_y_continuous(labels = percent)

by_class <- items %>% 
  drop_na(frequency, class) %>% 
  ggplot(aes(frequency, lv, fill = class, colour = class)) +
  facet_wrap(vars(class), ncol = 2) +
  geom_point(alpha = 0.25, shape = 1, stroke = 1, show.legend = FALSE) +
  labs(y = "Levenshtein similarity", x = "Lexical frequency", colour = "cognateness", fill = "Cognateness") +
  scale_y_continuous(labels = percent)

(by_language + by_class) +
  plot_layout(2) &
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank()
  )

@@

@@

```{r participants_time, message=FALSE, warning=FALSE, paged.print=FALSE}
participants %>% 
  group_by(time_stamp, dominant_language) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  group_by(dominant_language) %>% 
  mutate(n = cumsum(n)) %>% 
  ggplot(aes(time_stamp, n, colour = dominant_language, fill = dominant_language)) +
  geom_line(size = 1) +
  labs(x = "Date", y = "Number of responses", colour = "Dominant language") +
  scale_y_continuous(breaks = seq(0, 250, 50)) +
  guides(colour = guide_legend(ncol = 2)) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    axis.title.x = element_blank(),
  )

@@

```{r participants_age, message=FALSE, warning=FALSE, paged.print=FALSE}

participants %>% 
  mutate(age = round(age)) %>% 
  count(lp, dominant_language, age) %>% 
  ggplot(aes(as.factor(age), n, fill = interaction(lp, dominant_language, sep = " - "))) +
  geom_col(position = position_fill(), size = 0.5, colour = "white") +
  guides(colour = guide_legend(ncol = 2)) +
  coord_flip() +
  labs(y = "Proportion of the sample", x = "Age (months)", colour = "LP - Dominant language") +
  scale_y_continuous(labels = scales::percent) +
  theme(
    axis.text.y = element_text(size = 9),
    legend.position = "top",
    legend.title = element_blank(),
  ) +

  participants %>% 
  mutate(age = round(age)) %>% 
  count(lp, dominant_language, age) %>% 
  ggplot(aes(as.factor(age), n, fill = interaction(lp, dominant_language, sep = " - "))) +
  geom_col(size = 0.5, colour = "white") +
  labs(y = "Number of participants", x = "Age (months)") +
  guides(colour = guide_legend(ncol = 2)) +
  coord_flip() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()

  ) +

  plot_layout(nrow = 1, guides = "collect") &
  theme(
    legend.position = "top", 
    panel.grid.major = element_blank()
  )

@@

```{r participants_lp, message=FALSE, warning=FALSE, paged.print=FALSE}
labs <- c("0-10%", "10-20%", "20-30%", "30-40%", "40-50%")

tbl <- participants %>% 
  mutate(
    age = round(age),
    doe_2 = ifelse(dominant_language=="Catalan", doe_spanish, doe_catalan),
    doe_2 = cut(doe_2, breaks = seq(0, 0.5, 0.1), include.lowest = TRUE, labels = labs)
  ) %>% 
  count(age, doe_2) %>% 
  arrange(doe_2) %>% 
  mutate(doe_2 = fct_inorder(doe_2)) %>% 
  pivot_wider(names_from = doe_2, values_from = n, values_fill = 0)

gt(tbl) %>% 
  tab_spanner(label = "Number of participants by DOE", columns = 2:6) %>% 
  data_color(
    columns = 2:6,
    colors = scales::col_numeric(
      palette = c("white", "orange"),
      domain = c(0, max(tbl[,2:ncol(tbl)]))
    )
  ) %>% 
  cols_label(
    age = md("**Age<br>(months)**")
  ) %>% 
  as_raw_html()

@@

```{r participants_edu, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
tbl <- participants %>%
  mutate(edu_parent = fct_explicit_na(as.factor(edu_parent))) %>% 
  count(lp, dominant_language, edu_parent) %>% 
  right_join(
    expand_grid(
      lp = c("Monolingual", "Bilingual"),
      dominant_language = c("Catalan", "Spanish"),
      edu_parent = unique(participants$edu_parent)
    )
  ) %>% 
  replace_na(list(n = 0)) %>% 
  rename(total = n) %>% 
  pivot_wider(names_from = edu_parent, values_from = total, values_fill = 0) %>%
  mutate(total = rowSums(cbind(.[,3:8]))) %>% 
  rename(Total = total) %>% 
  select(-"NA")

gt(tbl, groupname_col = "lp") %>% 
  cols_label(
    lp = "Lang. Profile",
    dominant_language = "L1"
  ) %>% 
  summary_rows(
    columns = 3:8, 
    fns = list(Total = ~sum(.)),
    formatter = fmt_number,
    decimals = 0
  ) %>% 
  tab_style(
    style = list(cell_borders(weight = px(2), sides = "left")),
    locations = cells_body(columns = 10, rows = 1:4)
  ) %>% 
  as_raw_html()
```

@@

We initially fitted a null model (`fit_0`) than only included the predictors `age` and `frequency` as nuisance parameters, along with random intercepts by `id` and `item`, and random slopes of `frequency` by `id`, and `age_center` by `item`, and their correlation parameter. We then expanded this model (`fit_1`) to include the main effect of `doe`, and the `doe` by `item` random slope. Finally, we added the main effect `cognate` (`fit_2`), its interaction with `doe` (`doe:cognate`), and  random slopes for `cognate` by `id`. The models were implemented in `brms` as:

* `understands/produces ~ 1 + age_center + frequency + (1 + age_center | item) + (1 + frequency | id)`
* `understands/produces ~ 1 + age_center + frequency + doe + (1 + age_center + doe | item) + (1 + frequency | id)`
* `understands/produces ~ 1 + age_center + frequency + doe*cognate + (1 + age_center + doe | item) + (1 + frequency + cognate| id)`

@@

## R code (`brms`)

```
understands ~
1 + age_center + frequency + lp*cognate +
(1 + age_center + lp | te) +
(1 + frequency + cognate | id),
family =  bernoulli("logit")
```

## Stan code

@@

// generated with brms 2.15.0

@@

@@

@@

int Kc = K - 1;
matrix[N, Kc] Xc;  // centered version of X without an intercept

@@

for (i in 2:K) {
means_X[i - 1] = mean(X[, i]);
Xc[, i - 1] = X[, i] - means_X[i - 1];

@@

real Intercept;  // temporary intercept for centered predictors

@@

@@

sd_1 = rep_vector(1, rows(sd_1));

@@

vector[N] mu = Intercept + rep_vector(0.0, N);

@@

mu[n] += r_1_1[J_1[n]] * Z_1_1[n] + r_1_2[J_1[n]] * Z_1_2[n] + r_1_3[J_1[n]] * Z_1_3[n] + r_2_1[J_2[n]] * Z_2_1[n] + r_2_2[J_2[n]] * Z_2_2[n] + r_2_3[J_2[n]] * Z_2_3[n];

@@

target += bernoulli_logit_glm_lpmf(Y | Xc, mu, b);

@@

target += normal_lpdf(b | 0, 3);
target += normal_lpdf(Intercept | 0, 3);

@@

target += lkj_corr_cholesky_lpdf(L_1 | 5);
target += normal_lpdf(sd_2 | 0, 0.5)
- 3 * normal_lccdf(0 | 0, 0.5);

@@

target += lkj_corr_cholesky_lpdf(L_2 | 5);

@@

// actual population-level intercept
real b_Intercept = Intercept - dot_product(means_X, b);

@@

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    message = TRUE,
    echo = FALSE,
    warning = TRUE,
    include = TRUE,
    out.width = "80%",
    dpi = 250
)
options(
    knitr.kable.NA = "-",
    knitr.duplicate.label = "allow",
    ggplot2.discrete.fill = ggsci::pal_d3()(4),
    ggplot2.discrete.colour = ggsci::pal_d3()(4),
    ggplot2.continuous.fill = ggplot2::scale_color_gradient,
    ggplot2.continuous.colour = ggplot2::scale_color_gradient
)

@@

The questionnaire was implemented on-line using the formR platform [@arslan2020formr], and was structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items (`r n_item_cognate["catalan-cognate"]` cognates, `r n_item_cognate["catalan-noncognate"]` non-cognates) and the Spanish inventory contained `r n_item_language["spanish"]` (`r n_item_cognate["spanish-cognate"]` cognates, `r n_item_cognate["spanish-noncognate"]` non-cognates). Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* [dog] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. When there were two acceptable translation equivalents for a given word, we included both in separate items (e.g., Catalan *acabar* [*to finish*] and Spanish *acabar* and *terminar*), or merged them into a single items (e.g., Spanish *mono* [*monkey*] and Catalan *mono/mico*. We included items from a diverse sample of `r n_categories` semantic/functional categories (see Appendix 1). For the analyses included in this study, we excluded items from the adverbs, auxiliary words, connectives, interjections and games and routines categories, so that only data from content words (nouns, adjectives, and verbs) were used.

For each word in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Some families filled a long version of the vocabulary checklists (800 translation equivalents; 800 items in Catalan, 800 items in Spanish), while others filled a shorter version (~400 translation equivalents, ~400 items in Catalan, ~400 items in Spanish). These last families were randomly allocated into one of four different subsets of the complete list of items. These lists were carefully designed so that each contained a representative subsample of the items from the complete list. Semantic/functional categories with less than 16 items--thus resulting in less than four items after dividing it in four lists--were not divided in the short version of the questionnaire: all of their items were included in the four lists. Another subset of items that were part of the trial lists of some experiments in the lab were also included in all versions. Table 2 in Appendix 1 shows the distribution of items across questionnaire versions. We excluded from the analysis multi-word items (e.g., *barrita de cereales* [cereal bar]) and items that included more than one word-form (e.g., *mono / mico*). Table 3 shows the classification of items in cognates and non-cognates and their lexical frequency scores across the four lists of the inventories.

@@

```{r items_frequency, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=7, fig.cap="Lexical frequencies. Mean, standard error, and 95\\% confidence interval of lexical frequencies of items included in the Catalan and Spanish lists, reported separately for identical cognates, non-identical cognates, and non-cognates."}

items %>%
    left_join(select(multilex_data$pool, te, language)) %>% 
    unnest(list) %>% 
    group_by(language, list) %>% 
    summarise_at(
        vars(freq, n_phon, lv), 
        lst(n = ~sum(!is.na(.)), mean, sd, min, max)
    ) %>% 
    select(-c(n_phon_n, lv_n)) %>% 
    rename(n = freq_n) %>% 
    gt(groupname_col = "language") %>% 
    fmt_number(matches("freq_|n_phon_mean|n_phon_sd")) %>% 
    fmt_percent(matches("lv_")) %>% 
    tab_spanner("Frequency (Zipf)", columns = starts_with("freq_")) %>% 
    tab_spanner("# Phonemes", columns = starts_with("n_phon_")) %>% 
    tab_spanner("Levenshtein", columns = starts_with("lv_")) %>% 
    cols_merge_range(col_begin = "freq_min", col_end = "freq_max") %>% 
    cols_merge_range(col_begin = "n_phon_min", col_end = "n_phon_max") %>% 
    cols_merge_range(col_begin = "lv_min", col_end = "lv_max") %>% 
    cols_label(
        list = "",
        n = md("*N*"),
        freq_mean = md("Mean"),
        freq_sd = md("*SD*"),
        freq_min = md("Range"),
        n_phon_mean = md("Mean"),
        n_phon_sd = md("*SD*"),
        n_phon_min = md("Range"),
        lv_mean = md("Mean"),
        lv_sd = md("*SD*"),
        lv_min = md("Range")
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(everything())
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(everything())
    )

@@

```{r items-frequency, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

items %>% 
    mutate(language = ifelse(str_detect(item, "cat_"), "Catalan", "Spanish")) %>% 
    ggplot() +
    aes(
        x = freq,
        fill = language,
        colour = language,
        shape = language
    ) +
    stat_slab(
        slab_colour = "white",
        slab_size = 1,
        alpha = 0.5,
    ) +
    stat_pointinterval(
        aes(y = -0.1),
        position = position_dodge(width = 0.1),
        point_size = 3
    ) +
    geom_hline(yintercept = 0, colour = "black", size = 0.5) +
    labs(
        x = "Frequency (Zipf)",
        y = "Probability density",
        fill = "Language",
        shape = "Language",
        colour = "Language"
    ) +
    scale_y_continuous(labels = percent,  limits = c(-0.15, 1), breaks = seq(0, 1, 0.25)) +
    theme(
        legend.position = "top",
        axis.title = element_blank()
    )

@@

## # Phonemes

```{r items-nphon, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

items %>%
    mutate(language = ifelse(str_detect(item, "cat_"), "Catalan", "Spanish")) %>% 
    count(language, n_phon) %>% 
    mutate(prop = n/sum(.$n)) %>% 
    ggplot() +
    aes(
        x = n_phon,
        y = prop,
        fill = language,
        colour = language,
        shape = language
    ) +
    geom_col(
        position = position_dodge()
    ) +
    geom_hline(yintercept = 0, colour = "black", size = 0.5) +
    labs(
        x = "Frequency (Zipf)",
        y = "Probability density",
        fill = "Language",
        shape = "Language",
        colour = "Language"
    ) +
    scale_y_continuous(labels = percent, breaks = seq(0, 1, 0.05)) +
    theme(
        legend.position = "top",
        axis.title = element_blank()
    )

## Levenshtein

```{r items-lv, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

items %>% 
    distinct(te, .keep_all = TRUE) %>% 
    mutate(lv_cut = cut(lv, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>% 
    count(lv_cut) %>% 
    mutate(prop = n/sum(.$n)) %>% 
    ggplot() +
    aes(x = lv_cut, y = prop) +
    geom_col(fill = ggsci::pal_d3()(4)[1]) +
    labs(
        x = "Levenshtein",
        y = "Probability density",
    ) +
    theme(
        legend.position = "top",
        axis.title = element_blank()
    )

## Frequency by # Phonemes

```{r items-frequency-phonemes, message=FALSE, warning=FALSE, paged.print=FALSE}

items %>% 
    left_join(select(pool, item, language)) %>% 
    drop_na(freq) %>% 
    ggplot(aes(n_phon, freq, colour = language, fill = language)) +
    facet_wrap(~language) +
    geom_point(size = 1, alpha = 0.25) +
    geom_smooth(method = "lm", formula = "y ~ x") +
    labs(
        x = "# Phonemes", 
        y = "Lexical frequency (Zipf score)\nExtracted from SUBTLEX", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    )

## Frequency by Levenshtein

```{r items-frequency-levenshtein, message=FALSE, warning=FALSE, paged.print=FALSE}

items %>% 
    left_join(select(pool, item, language)) %>% 
    drop_na(freq) %>% 
    ggplot(aes(freq, lv, colour = language, fill = language)) +
    facet_wrap(~language) +
    geom_point(size = 1, alpha = 0.5) +
    geom_smooth(method = "lm", formula = "y ~ x") +
    labs(
        y = "Levenshtein similarity", 
        x = "Lexical frequency (Zipf score)\nExtracted from SUBTLEX", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    theme(
        legend.position = "none",
        panel.grid.major.y = element_blank()
    )

## # Phonemes by Levenshtein

```{r items-phonemes-lv, message=FALSE, warning=FALSE, paged.print=FALSE}
items %>% 
    left_join(select(pool, item, language)) %>% 
    drop_na(freq) %>% 
    ggplot(aes(n_phon, lv, colour = language, fill = language)) +
    facet_wrap(~language) +
    geom_point(size = 1, alpha = 0.25) +
    geom_smooth(method = "lm", formula = "y ~ x") +
    labs(
        x = "# Phonemes", 
        y = "Levenshtein similarity", 
        colour = "Language", 
        fill = "Language"
    ) +
    scale_y_continuous(labels = percent) +
    scale_x_continuous(breaks = seq(0, 16, 1)) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()
    ) 
```

@@

```{r participants_time, message=FALSE, warning=FALSE, paged.print=FALSE}
participants %>% 
    mutate(dominant_language = ifelse(doe_catalan >= doe_spanish, "Catalan", "Spanish")) %>% 
    group_by(time_stamp, dominant_language) %>% 
    summarise(n = n(), .groups = "drop") %>% 
    group_by(dominant_language) %>% 
    mutate(n = cumsum(n)) %>% 
    ggplot(aes(time_stamp, n, colour = dominant_language, fill = dominant_language)) +
    geom_vline(xintercept = ymd("2020-03-15"), size = 0.5) +
    annotate(geom = "text", y = 350, x = ymd("2020-03-15"), label = "COVID-19 lockdown",
             angle = 90, vjust = 1.5, hjust = 1) +
    geom_line(size = 1) +
    labs(x = "Date", y = "Number of responses", colour = "Dominant language") +
    scale_y_continuous(breaks = seq(0, 400, 50), limits = c(0, 400)) +
    guides(colour = guide_legend(ncol = 2)) +
    theme(
        legend.position = "top",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
    )

@@

```{r participants-age, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
participants %>% 
    mutate(age = floor(age)) %>% 
    count(age) %>% 
    ggplot() +
    aes(x = age, y = n) +
    geom_col() +
    geom_text(aes(label = n), size = 3.5, vjust = -1) +
    labs(
        x = "Age (months)",
        y = "# participants"
    ) +
    scale_x_continuous(breaks = seq(0, 40, 2)) +
    scale_y_continuous(limits = c(0, 55), breaks = seq(0, 55, 10)) +
    theme(
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()

    )
```

@@

```{r participants-lp, message=FALSE, warning=FALSE, paged.print=FALSE}

participants %>% 
    mutate(
        dominant_language = ifelse(doe_catalan >= doe_spanish, "Catalan", "Spanish"),
        age = round(age),
        id = paste(id, "(", time, ")"),
        rank = rank(doe_catalan)
    ) %>% 
    pivot_longer(
        starts_with("doe_"), 
        names_to = "language",
        values_to = "doe",
        names_prefix = "doe_",
        names_transform = list(language = str_to_sentence)
    ) %>% 
    ggplot() +
    aes(x = doe, y = reorder(id, -rank), fill = language) +
    geom_col(width = 0.75, position = position_fill(), colour = "white", size = 0.1) +
    labs(
        x = "Degree of exposure to Catalan",
        y = "Participant",
        colour = "Language",
        fill = "Language"
    ) +
    scale_x_continuous(labels = percent) +
    theme(
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank()
    )

@@

```{r participants_edu, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
participants %>%
    left_join(select(multilex_data$logs, id, time, edu_parent)) %>% 
    mutate(
        edu_parent = fct_explicit_na(as.factor(edu_parent)),
        doe_catalan = cut(doe_catalan, breaks = seq(0, 1, 0.1), include.lowest = TRUE)
    ) %>% 
    count(doe_catalan, edu_parent) %>% 
    right_join(
        expand_grid(
            doe_catalan = levels(.$doe_catalan),
            edu_parent = unique(multilex_data$logs$edu_parent)
        )
    ) %>% 
    rename(total = n) %>% 
    pivot_wider(names_from = edu_parent, values_from = total, values_fill = 0) %>%
    mutate(total = rowSums(cbind(.[,3:8]))) %>% 
    rename(Total = total) %>% 
    select(-"NA") %>% 
    gt() %>% 
    cols_label(
        doe_catalan = "DoE Catalan"
    ) %>% 
    summary_rows(
        columns = 2:7, 
        fns = list(Total = ~sum(., na.rm = TRUE)),
        formatter = fmt_number,
        decimals = 0
    ) %>% 
    fmt_missing(columns = 3:8, missing_text = "--") %>% 
    tab_spanner("Educational attainment", 3:8) %>% 
    tab_style(
        style = list(cell_borders(weight = px(2), sides = "left", color = "grey50")),
        locations = cells_body(columns = 9, rows = 1:10)
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(columns = 2:7)
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_labels(columns = c(1, 8))
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(spanners = "Educational attainment")
    )
```

@@

We initially fitted a null model (`fit_0`) than only included the predictors `age` and `frequency` as nuisance parameters, along with random intercepts by `id` and `item`, and random slopes of `frequency` by `id`, and `age_std` by `item`, and their correlation parameter. We then expanded this model (`fit_1`) to include the main effect of `doe`, and the `doe` by `item` random slope. Finally, we added the main effect `cognate` (`fit_2`), its interaction with `doe` (`doe:cognate`), and  random slopes for `cognate` by `id`. The models were implemented in `brms` as:

* `understands/produces ~ 1 + age_std + frequency + (1 + age_std | item) + (1 + frequency | id)`
* `understands/produces ~ 1 + age_std + frequency + doe + (1 + age_std + doe | item) + (1 + frequency | id)`
* `understands/produces ~ 1 + age_std + frequency + doe*cognate + (1 + age_std + doe | item) + (1 + frequency + cognate| id)`

@@

## R code (`brms`) {.tabset}

### Model 0

```{r}
print(model_fit_0$formula)
```

### Model 1

```{r}
print(model_fit_1$formula)
```

### Model 2

```{r}
print(model_fit_2$formula)
```

### Model 3

```{r}
print(model_fit_3$formula)
```

### Model 4

```{r}
print(model_fit_4$formula)
```

### Model 5

```{r}
print(model_fit_5$formula)
```

### Model 6

```{r}
print(model_fit_6$formula)
```

### Model 7

```{r}
print(model_fit_7$formula)
```

## Stan code {.tabset}

@@

### Model 0

```{stan, eval = FALSE, tidy = FALSE}
// generated with brms 2.16.1
functions {
  /* cratio-logit log-PDF for a single response
   * Args:
   *   y: response category
   *   mu: latent mean parameter
   *   disc: discrimination parameter
   *   thres: ordinal thresholds
   * Returns:
   *   a scalar to be added to the log posterior
   */
   real cratio_logit_lpmf(int y, real mu, real disc, vector thres) {
     int nthres = num_elements(thres);
     vector[nthres + 1] p;
     vector[nthres] q;
     int k = 1;
     while (k <= min(y, nthres)) {
       q[k] = log_inv_logit(disc * (mu - thres[k]));
       p[k] = log1m_exp(q[k]);
       for (kk in 1:(k - 1)) p[k] = p[k] + q[kk];
       k += 1;
     }
     if (y == nthres + 1) {
       p[nthres + 1] = sum(q);
     }
     return p[y];
   }
  /* cratio-logit log-PDF for a single response and merged thresholds
   * Args:
   *   y: response category
   *   mu: latent mean parameter
   *   disc: discrimination parameter
   *   thres: vector of merged ordinal thresholds
   *   j: start and end index for the applid threshold within 'thres'
   * Returns:
   *   a scalar to be added to the log posterior
   */
   real cratio_logit_merged_lpmf(int y, real mu, real disc, vector thres, int[] j) {
     return cratio_logit_lpmf(y | mu, disc, thres[j[1]:j[2]]);
   }
}
data {
  int<lower=1> N;  // total number of observations
  int Y[N];  // response variable
  int<lower=2> nthres;  // number of thresholds
  // data for group-level effects of ID 1
  int<lower=1> N_1;  // number of grouping levels
  int<lower=1> M_1;  // number of coefficients per level
  int<lower=1> J_1[N];  // grouping indicator per observation
  // group-level predictor values
  vector[N] Z_1_1;
  // data for group-level effects of ID 2
  int<lower=1> N_2;  // number of grouping levels
  int<lower=1> M_2;  // number of coefficients per level
  int<lower=1> J_2[N];  // grouping indicator per observation
  // group-level predictor values
  vector[N] Z_2_1;
  int prior_only;  // should the likelihood be ignored?
}
transformed data {
}
parameters {
  vector[nthres] Intercept;  // temporary thresholds for centered predictors
  vector<lower=0>[M_1] sd_1;  // group-level standard deviations
  vector[N_1] z_1[M_1];  // standardized group-level effects
  vector<lower=0>[M_2] sd_2;  // group-level standard deviations
  vector[N_2] z_2[M_2];  // standardized group-level effects
}
transformed parameters {
  real<lower=0> disc = 1;  // discrimination parameters
  vector[N_1] r_1_1;  // actual group-level effects
  vector[N_2] r_2_1;  // actual group-level effects
  r_1_1 = (sd_1[1] * (z_1[1]));
  r_2_1 = (sd_2[1] * (z_2[1]));
}
model {
  // likelihood including constants
  if (!prior_only) {
    // initialize linear predictor term
    vector[N] mu = rep_vector(0.0, N);
    for (n in 1:N) {
      // add more terms to the linear predictor
      mu[n] += r_1_1[J_1[n]] * Z_1_1[n] + r_2_1[J_2[n]] * Z_2_1[n];
    }
    for (n in 1:N) {
      target += cratio_logit_lpmf(Y[n] | mu[n], disc, Intercept);
    }
  }
  // priors including constants
  target += normal_lpdf(Intercept | -0.25, 0.1);
  target += normal_lpdf(sd_1 | 1, 0.1)
    - 1 * normal_lccdf(0 | 1, 0.1);
  target += std_normal_lpdf(z_1[1]);
  target += normal_lpdf(sd_2 | 1, 0.1)
    - 1 * normal_lccdf(0 | 1, 0.1);
  target += std_normal_lpdf(z_2[1]);
}
generated quantities {
  // compute actual thresholds
  vector[nthres] b_Intercept = Intercept;
  // additionally sample draws from priors
  real prior_Intercept = normal_rng(-0.25,0.1);
  real prior_sd_1 = normal_rng(1,0.1);
  real prior_sd_2 = normal_rng(1,0.1);
  // use rejection sampling for truncated priors
  while (prior_sd_1 < 0) {
    prior_sd_1 = normal_rng(1,0.1);
  }
  while (prior_sd_2 < 0) {
    prior_sd_2 = normal_rng(1,0.1);
  }
}
```

### Model 1

```{stan, eval = FALSE, tidy = FALSE}
// generated with brms 2.16.1

@@

  /* cratio-logit log-PDF for a single response
   *   y: response category
   *   mu: latent mean parameter
   *   disc: discrimination parameter
   *   thres: ordinal thresholds
   *   a scalar to be added to the log posterior
   real cratio_logit_lpmf(int y, real mu, real disc, vector thres) {
     int nthres = num_elements(thres);
     vector[nthres + 1] p;
     vector[nthres] q;
     int k = 1;
     while (k <= min(y, nthres)) {
       q[k] = log_inv_logit(disc * (mu - thres[k]));
       p[k] = log1m_exp(q[k]);
       for (kk in 1:(k - 1)) p[k] = p[k] + q[kk];
       k += 1;
     if (y == nthres + 1) {
       p[nthres + 1] = sum(q);
     return p[y];
  /* cratio-logit log-PDF for a single response and merged thresholds
   *   y: response category
   *   mu: latent mean parameter
   *   disc: discrimination parameter
   *   thres: vector of merged ordinal thresholds
   *   j: start and end index for the applid threshold within 'thres'
   *   a scalar to be added to the log posterior
   real cratio_logit_merged_lpmf(int y, real mu, real disc, vector thres, int[] j) {
     return cratio_logit_lpmf(y | mu, disc, thres[j[1]:j[2]]);

@@

  int<lower=2> nthres;  // number of thresholds
  int Kc = K;
  matrix[N, Kc] Xc;  // centered version of X
  for (i in 1:K) {
    means_X[i] = mean(X[, i]);
    Xc[, i] = X[, i] - means_X[i];
  vector[nthres] Intercept;  // temporary thresholds for centered predictors
  real<lower=0> disc = 1;  // discrimination parameters
    vector[N] mu = Xc * b;
      mu[n] += r_1_1[J_1[n]] * Z_1_1[n] + r_1_2[J_1[n]] * Z_1_2[n] + r_2_1[J_2[n]] * Z_2_1[n] + r_2_2[J_2[n]] * Z_2_2[n];
      target += cratio_logit_lpmf(Y[n] | mu[n], disc, Intercept);
  target += normal_lpdf(b[1] | 1, 0.1);
  target += normal_lpdf(Intercept | -0.25, 0.1);
  target += normal_lpdf(sd_1 | 1, 0.1)
    - 2 * normal_lccdf(0 | 1, 0.1);
  target += lkj_corr_cholesky_lpdf(L_1 | 2);
  target += normal_lpdf(sd_2 | 1, 0.1)
    - 2 * normal_lccdf(0 | 1, 0.1);
  target += lkj_corr_cholesky_lpdf(L_2 | 2);
  // compute actual thresholds
  vector[nthres] b_Intercept = Intercept + dot_product(means_X, b);
  // additionally sample draws from priors
  real prior_b_1 = normal_rng(1,0.1);
  real prior_Intercept = normal_rng(-0.25,0.1);
  real prior_sd_1 = normal_rng(1,0.1);
  real prior_cor_1 = lkj_corr_rng(M_1,2)[1, 2];
  real prior_sd_2 = normal_rng(1,0.1);
  real prior_cor_2 = lkj_corr_rng(M_2,2)[1, 2];
  // use rejection sampling for truncated priors
  while (prior_sd_1 < 0) {
    prior_sd_1 = normal_rng(1,0.1);
  while (prior_sd_2 < 0) {
    prior_sd_2 = normal_rng(1,0.1);

### Model 2

```{stan, eval = FALSE, tidy = FALSE}
// generated with brms 2.16.1
  /* cratio-logit log-PDF for a single response
   *   y: response category
   *   mu: latent mean parameter
   *   disc: discrimination parameter
   *   thres: ordinal thresholds
   *   a scalar to be added to the log posterior
   real cratio_logit_lpmf(int y, real mu, real disc, vector thres) {
     int nthres = num_elements(thres);
     vector[nthres + 1] p;
     vector[nthres] q;
     int k = 1;
     while (k <= min(y, nthres)) {
       q[k] = log_inv_logit(disc * (mu - thres[k]));
       p[k] = log1m_exp(q[k]);
       for (kk in 1:(k - 1)) p[k] = p[k] + q[kk];
       k += 1;
     if (y == nthres + 1) {
       p[nthres + 1] = sum(q);
     return p[y];
  /* cratio-logit log-PDF for a single response and merged thresholds
   *   y: response category
   *   mu: latent mean parameter
   *   disc: discrimination parameter
   *   thres: vector of merged ordinal thresholds
   *   j: start and end index for the applid threshold within 'thres'
   *   a scalar to be added to the log posterior
   real cratio_logit_merged_lpmf(int y, real mu, real disc, vector thres, int[] j) {
     return cratio_logit_lpmf(y | mu, disc, thres[j[1]:j[2]]);
  int<lower=2> nthres;  // number of thresholds

@@

  int Kc = K;
  matrix[N, Kc] Xc;  // centered version of X

@@

  for (i in 1:K) {
    means_X[i] = mean(X[, i]);
    Xc[, i] = X[, i] - means_X[i];

@@

  vector[nthres] Intercept;  // temporary thresholds for centered predictors

@@

  real<lower=0> disc = 1;  // discrimination parameters

@@

@@

    vector[N] mu = Xc * b;

@@

      mu[n] += r_1_1[J_1[n]] * Z_1_1[n] + r_1_2[J_1[n]] * Z_1_2[n] + r_1_3[J_1[n]] * Z_1_3[n] + r_2_1[J_2[n]] * Z_2_1[n] + r_2_2[J_2[n]] * Z_2_2[n];

@@

      target += cratio_logit_lpmf(Y[n] | mu[n], disc, Intercept);

@@

  target += normal_lpdf(b[1] | 1, 0.1);
  target += normal_lpdf(b[2] | 0, 0.1);
  target += normal_lpdf(Intercept | -0.25, 0.1);
  target += normal_lpdf(sd_1 | 1, 0.1)
    - 3 * normal_lccdf(0 | 1, 0.1);

@@

  target += lkj_corr_cholesky_lpdf(L_1 | 2);
  target += normal_lpdf(sd_2 | 1, 0.1)
    - 2 * normal_lccdf(0 | 1, 0.1);

@@

  target += lkj_corr_cholesky_lpdf(L_2 | 2);

@@

  // compute actual thresholds
  vector[nthres] b_Intercept = Intercept + dot_product(means_X, b);

@@

  // additionally sample draws from priors
  real prior_b_1 = normal_rng(1,0.1);
  real prior_b_2 = normal_rng(0,0.1);
  real prior_Intercept = normal_rng(-0.25,0.1);
  real prior_sd_1 = normal_rng(1,0.1);
  real prior_cor_1 = lkj_corr_rng(M_1,2)[1, 2];
  real prior_sd_2 = normal_rng(1,0.1);
  real prior_cor_2 = lkj_corr_rng(M_2,2)[1, 2];

@@

  // use rejection sampling for truncated priors
  while (prior_sd_1 < 0) {
    prior_sd_1 = normal_rng(1,0.1);
  while (prior_sd_2 < 0) {
    prior_sd_2 = normal_rng(1,0.1);
